Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2023-05-12 noise_covid_v2.Rmd
scripts for visualization

# ```{r}
# disp_stats <- c()
# for (i in 1:dim(new_cases_subset[state_index])[1]){
#   Y <- new_cases_subset[i,]
#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
# }
# ```

@@

```{r}
source("code/squared.res_v_squared.fitted.R")
```

@@

We tested the hypothesis that the incidence dispersion parameter decreases at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$. Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020. The Wald test is used, where the estimates of $\theta$ in different segments are considered independent. This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information.

@@

```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
source("code/wald_counties.R")
```
```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
source("code/LRT_counties.R")
```

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_Wald.R")
```

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.

The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean. As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_LRT.R")
```

>>>>

@@

```{r}
par(mfrow = c(1,2))
state <- "NY"
source("code/squared.res_v_squared.fitted.R")

state <- "FL"
source("code/squared.res_v_squared.fitted.R")
```

@@

We tested the hypothesis that the incidence dispersion parameter decreases at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$. Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020. The Wald test is used, where the estimates of $\theta$ in different segments are considered independent. This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information matrix.

@@

```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
par(mfrow = c(1,2))
state <- "NY"
source("code/wald_counties.R")

state <- "FL"
source("code/wald_counties.R")
```
```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
par(mfrow = c(1,2))
state <- "NY"
source("code/lrt_counties.R")
NY_sig <- (which(pvals < 0.05))

state <- "FL"
source("code/lrt_counties.R")
FL_sig <- print(which(pvals < 0.05))

```

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_Wald.R")
```
To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean. As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_LRT.R")
```
There appears to be slight Type I error inflation in both tests.
